### 2024-05-17 Manuscript/main.tex
Update main.tex

In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}.
Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any observable characteristic (e.g. morphological, anatomical, ecological, physiological, behavioural, phenological, etc.) measured on individual organisms at any level, from genes to whole organisms.
A trait-based approach has two advantages for answering core questions in ecology and evolution.
First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms’ traits at the centre of natural selection rather than organisms themselves.
Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g. unrelated species pools) and time (e.g. anatomical traits are comparable between palaeontology and ecology, which is not always the case with species; \citealt{luza2023going}).
Already used routinely in palaeontology \citep{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \citep{mcgill2006rebuilding}, biogeography \citep{violle2014emergence}, conservation biology \citep{chichorro2022trait}, micro- \citep{chapin1993evolution}, macro-evolution \citep{guillerme2023innovation} and applied fields (e.g. agronomy \citealt{martin2015plant}).
This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism) (see Box 1).

@@

While the first part of this framework (``what?'') has been studied extensively in the last decade this was often undertaken outside of the ``what, how and why'' framework.
Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
Using an inappropriate metric for describing a pattern can lead to biased conclusions.
For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home two both populations and another one home to only one of the population (the process - ``how?'').
In such a scenario, measuring the occupied area (i.e. $km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations.
In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin’s finches share resources depending on the existence or not of competition.

@@

\item Decreasing size: by removing the required amount (20\%, 40\%, 60\% or 80\%) of data away from a distance (radius)  $\rho$ of the centre of the trait-space.
This approximates our \textbf{equalising mechanism}.

@@

    \end{tabular}
    \caption{\scriptsize{\textbf{Description of the stressors applied to the simulated data and the mechanism each approximated.} Here we distinguish between the biological mechanism we are trying to simulate and the algorithmic one we used to simulate it. We also provide a more detailed description of the stressor. Note that in biological data, we don’t expect any of the mechanisms to be applied to the data alone (e.g. equalising fitness and facilitation can both act on the distribution of species traits). Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution). These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
}
    \label{Tab:mechanisms}
\end{table}

@@

To ease interpretation of the results, w e centred the results for the 14 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null and 1 and –1 the maximum positive and negative deviations relative to the null.
In parallel, we also compared the distribution of the 200 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
The results of these tests are displayed in supplementary tables 1 to 3 and in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).

@@

Finally, we measured the effect of the different levels of data removal (20\%, 40\%, 60\% and 80\%) by fitting a linear regression using the model centred and scaled metric score ~ amount of data removed.
For each of these models, we reported the model fit (adjusted R2) and whether the slope was clearly distinguishable from 0 (p-values). See figure 2.

@@

We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian Islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers \citep{Walther2022}.
A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE) and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain).
We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from \cite{sayol2021loss} data.
We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}
We used 9 morphological traits associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; kipp’s distance (length from the first secondary feather to the tip of the longest primary feather; \citealt{tobias2022avonet}); and tail length
Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.

The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years.
Fifty-five species are known to have gone extinct prior to 1500 CE and 26 after 1500 CE.
For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a trait space richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a trait space richness reduction of 69\%).

To build the community trait-space, we undertook a PCA including all 118 species, which we then subsetted to calculate the trait-space for the historic and extant datasets.
We selected the first 5 axes to represent the trait-spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
We applied the same procedure as for the simulated data by simulating a null mechanism to compare to the observed ones by randomly removing 55 and 81 species for each stressor and scaling the results proportionally to this null mechanism (as described above).

@@

We tested 14 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
Our results show that, unexpectedly, different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
Therefore, the choice of traits space occupancy metric is essential to accurately describing a pattern of interest.

>>>>

In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}.
Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any observable characteristic (e.g. morphological, anatomical, ecological, physiological, behavioural, phenological, etc.) measured on individual organisms at any level, from genes to whole organisms.
A trait-based approach has two advantages for answering core questions in ecology and evolution.
First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms' traits at the centre of natural selection rather than organisms themselves.
Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g. unrelated species pools) and time (e.g. anatomical traits are comparable between palaeontology and ecology, which is not always the case with species; \citealt{luza2023going}).
Already used routinely in palaeontology \citep{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \citep{mcgill2006rebuilding}, biogeography \citep{violle2014emergence}, conservation biology \citep{chichorro2022trait}, micro- \citep{chapin1993evolution}, macro-evolution \citep{guillerme2023innovation} and applied fields (e.g. agronomy \citealt{martin2015plant}).
This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism) (see Box 1).

@@

While the first part of this framework (``what?'') has been studied extensively in the last decade this was often undertaken outside of the ``what, how and why'' framework.
Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
Using an inappropriate metric for describing a pattern can lead to biased conclusions.
For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home two both populations and another one home to only one of the population (the process - ``how?'').
In such a scenario, measuring the occupied area (i.e. $km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations.
In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin's finches share resources depending on the existence or not of competition.

@@

\item Decreasing size: by removing the required amount (20\%, 40\%, 60\% or 80\%) of data away from a distance (radius)  $\rho$ of the centre of the trait space.
This approximates our \textbf{equalising mechanism}.

@@

    \end{tabular}
    \caption{\scriptsize{\textbf{Description of the stressors applied to the simulated data and the mechanism each approximated.} Here we distinguish between the biological mechanism we are trying to simulate and the algorithmic one we used to simulate it. We also provide a more detailed description of the stressor. Note that in biological data, we don't expect any of the mechanisms to be applied to the data alone (e.g. equalising fitness and facilitation can both act on the distribution of species traits). Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution). These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
}
    \label{Tab:mechanisms}
\end{table}

@@

To ease interpretation of the results, we centred the results for the 12 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null and 1 and –1 the maximum positive and negative deviations relative to the null.
In parallel, we also compared the distribution of the 171 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
The results of these tests are displayed in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).

@@

Finally, we measured the effect of the different levels of data removal (20\%, 40\%, 60\% and 80\%) by fitting a linear regression using the model centred and scaled metric score ~ amount of data removed.
For each of these models, we reported the model fit (adjusted $R^2$) in the supplementary tables 1 to 3 and whether the slope was clearly distinguishable from 0 (Figure \ref{Fig:simulation_results}).

@@

We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers \citep{Walther2022}.
A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE) and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (\textit{Asio flammeus} was removed from the historic and extant species lists as its colonisation status is uncertain).
We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from \cite{sayol2021loss}.
We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}.
We used 9 morphological traits associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; the length from the first secondary feather to the tip of the longest primary feather (Kipp's distance); and tail length.
Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.

The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years.
55 species are known to have gone extinct prior to 1500 CE and 26 after 1500 CE.
For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a trait space richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a trait space richness reduction of 69\%).

To build the community trait space, we undertook a PCA including all 118 species, which we then subset to calculate the trait space for the historic and extant datasets.
We selected the first 5 axes to represent the trait spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
We applied the same procedure as for the simulated data by simulating a null mechanism to compare to the observed ones by randomly removing 55 and 81 species for each stressor and scaling the results proportionally to this null mechanism (as described above).

@@

We tested 12 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
Our results show that, unexpectedly, different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
Therefore, the choice of traits space occupancy metric is essential to accurately describing a pattern of interest.
